---
title: "High dimensional polytope sampling with samplelim"
author: "Philippe Regnault"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Polytope sampling with samplelim}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Vignettes are long form documentation commonly included in packages. Because they are part of the distribution of the package, they need to be as compact as possible. The `html_vignette` output type provides a custom style sheet (and tweaks some options) to ensure that the resulting html is as small as possible. The `html_vignette` format:

- Never uses retina figures
- Has a smaller default figure size
- Uses a custom CSS stylesheet instead of the default Twitter Bootstrap style

## Vignette Info

Note the various macros within the `vignette` section of the metadata block above. These are required in order to instruct R how to build the vignette. Note that you should change the `title` field and the `\VignetteIndexEntry` to match the title of your vignette.

## Styles

The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows:

    output: 
      rmarkdown::html_vignette:
        css: mystyles.css

## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))

## Describing the polytope to sample

A polytope $\mathcal{P}$ is an $n$-dimensional convex set defined as the intersection of hyper-planes and half-spaces. 
More precisely, $\mathcal{P} = \{ x \in \mathbb{R}^n: Ax = B, Gx \geq H \}$, 
where $A$ is an $m\times n$ matrix, with $m \leq n$, $B \in \mathbb{R}^m$, $G$ is a $k \times n$ matrix, with $k \geq ..$ and $H \in \mathbb{R}^k$. Inequality constraints, represented by the matrix $G$, involved in linear inverse models for metabolic networks make the polytope bounded.

The main argument `lim` of `rlim()` is a list with (at least) four components named `A`, `B`, `G` and `H`, the matrices and vectors defining the polytope to be sampled. Right below is a minimal example.

```{r minimalexample}
library("samplelim")
# Define equality and inequality constraints through matrices A, B, G, H
A <- matrix(rep(1, 3L), nrow = 1, ncol = 3)
B <- 1
G <- -matrix(c(0,0,1,
              0, 0, -1,
              0, 1, 0, 
              0, -1, 0,
              1, 0, 0, 
              -1, 0, 0),
            byrow = TRUE,
            nrow = 6, ncol = 3)
H <- -matrix(c(0.7, 0, 0.8, 0, 0.8, 0), nrow = 6)
# Store into a list
lim_exm <- list(A = A, B = B,G = G,H = H)
# Sampling into the polytope defined by these constraints
sample <- rlim(lim_exm, seed = 123)
# Show first points of the sample
head(sample)
```

In the previous -- pedagogical, example, we can visualize the sample generated by `rlim()` thanks to a 3D scatterplot.

```{r vizminexm, message = FALSE, eval = FALSE}
# Preparing sample for 3D scatterplot
colnames(sample) <- c("X", "Y","Z")
sample <- as.data.frame(sample)
library("plotly")
plot_sample <- plot_ly(sample, x = ~X, y = ~Y, z = ~Z, size = 0.2)
plot_sample <- plot_sample %>% add_markers()
plot_sample
```

